---
title: "Chapter 2: Linear Time Series Analysis and Its Applications"
author: "kongs"
date: "Sunday, July 06, 2014"
---

```{r, include=FALSE, cache = FALSE}
source("util/knitropts.R")
```

`Warning:`{.text-warning} Hyperlinks of the table of contents may be broken.
This hopefully will be fixed by Shinyapps.io later.

`Todo:`{.text-primary} Write a self-sufficient example 
for ARIMA estimation and diagnostic, using suggested packages.

`Todo:`{.text-primary} Compare different software working process, 
include `Mablab`, `Stata`, `Python`, `SAS`.

`Todo:`{.text-primary} Compare different software working process, 
include `Mablab`, `Stata`, `Python`, `SAS`.

# Stationary


# Correlation and Autocorrelation Function


## Autocorrelation Function (ACF)

$$ \rho_\ell
= \frac{\text{Cov}(r_t, r_{t-\ell})}{\sqrt{\text{Var}(r_t)\text{Var}(r_{t-\ell})}}
= \frac{\text{Cov}(r_t, r_{t-\ell})}{\text{Var}(r_t)}
= \frac{\gamma_\ell}{\gamma_0}
$$


## Test Individual ACF



## Portmanteau Test

Test jointly that several autocorrelations of $r_t$ are zero.

  * $H_0$: $\rho_1 = \rho_2 = ... = \rho_m = 0$
  
$$ Q(m) = T(T+2)\sum_{\ell=1}^{m} \frac{\hat{\rho}_{\ell}^2}{T - \ell} $$

```{r}
# library(fBasics)
# library(forecast)
source('util/AFTS_dataset.R')
```



```{r}
df = read.table(dataset$m_ibm3dx2608$url, header=T)
sibm = df$ibmrtn
Box.test(sibm, lag=12, type='Ljung')
sibm_acf = acf(sibm)
sibm_acf
sibm_pacf = pacf(sibm)
sibm_pacf

# If you want acf start with lag 1, as acf(lag=0) is always 1, you can give xlim starting from 1. Likely you can also set ylim to make it looks better.
acf(sibm, lag=12, xlim=c(1, 12))
```

`acf2{astsa}` is a handy function to plot both acf and pacf from $lag=1$,
sharing the same y axis range.

```{r}
acf_pacf = astsa::acf2(sibm)
head(acf_pacf)
```

Package `{fBasics}` also include another two functions that deal with correlations: `lacfPlot`, and `teffectPlot`.

```{r}
fBasics::lacfPlot(sibm)
fBasics::teffectPlot(sibm)
```


# White noise and linear time series

## Linear Time Series

$$ r_t = \mu + \sum_{i=0}^{\infty} \psi_i a_{t-i} $$

We have,

$$ E(r_t) = \mu, \text{Var}(r_t) = \sigma_a^2 \sum_{i=0}^\infty \psi_i^2 $$

The lag-$\ell$ autocovariance of $r_t$ is

$$ \gamma_\ell = \sigma_a^2 \sum_{j=0}^\infty \psi_j \psi_{j+\ell} $$

$$ \rho_\ell = \frac{\gamma_\ell}{\gamma_0} 
= \frac{\sum_{i=0}^\infty \psi_i \psi_{i+\ell}}{1+\sum_{i=1}^\infty \psi_i^2}
$$


# Simple AR Models

### An AR(1) Model

$$ r_t = \phi_0 + \phi_1 r_{t-1} + a_t$$
Where $\{a_t\}$ is assumed to be a white noise series 
with mean zero and variance $\sigma_a^2$.

$$ E(r_t|r_{t-1}) = \phi_0 + \phi_1 r_{t-1}, 
\text{Var}(r_t|r_{t-1}) = \text{Var}(a_t) = \sigma_a^2$$

### AR(p) Model

$$ r_t = \phi_0 + \phi_1 r_{t-1} + \ldots + \phi_p r_{t-p} + a_t $$

## Properties of AR Models

### AR(1) Model

$$ E(r_t) = \mu = \frac{\phi_0}{1-\phi_1} $$

$$ r_t - \mu = \phi_1(r_{t-1}-\mu) + a_t $$

$$ \text{Var}(r_t) = \frac{\sigma_a^2}{1-\phi_1^2} $$

$$ \rho_\ell = \phi_1 \rho_{\ell - 1} $$

### AR(2) Model

$$ (r_t - \mu) = \phi_1(r_{t-1} - \mu) + \phi_2(r_{t-2} - \mu) + a_t $$

$$ \rho_1 = \frac{\phi_1}{1-\phi_2} $$
$$ \rho_\ell = \phi_1 \rho_{\ell-1} + \phi_2 \rho_{\ell-2} $$

$$ (1-\phi_1 B - \phi_2 B^2)\phi_\ell = 0 $$

The solutions to the corresponding second-order polynomial equation are:

$$ x= \frac{\phi_1 \pm \sqrt{\phi_1^2 + 4 \phi_2}}{-2\phi_2} $$

$$ k = \frac{2\pi}{\cos^{-1}(\phi_1 / 2\sqrt{-\phi_2})}$$

If we write the solutions as $a \pm b i$, then we have 
$\phi_1 = 2a, \phi_2 = -(a^2 + b^2)$, and

$$ k = \frac{2\pi}{\cos^{-1}(a / \sqrt{a^2 + b^2})} $$

#### Example

```{r, strip.white=TRUE}
gnp = scan(file=dataset$dgnp82$url)
gnp_ts = ts(gnp, frequency=4, start=c(1947,2))
plot(gnp_ts)
points(gnp_ts, pch='*')

gnp_ar = ar(gnp, method='mle')
# Calculate $\phi_0$:
(1-sum(gnp_ar$ar)) * gnp_ar$x.mean
# Residual standard error:
sqrt(gnp_ar$var.pred)

# calculate business cycle k
p1 = c(1, -gnp_ar$ar)
roots = polyroot(p1)
roots
Mod(roots)
k = 2 * pi / acos(1.59023/1.913308)
k

# If using a general arima model with order (ar, 0, 0)
gnp_arima = arima(gnp, order=c(gnp_ar$order, 0, 0))
gnp_arima
coefs = gnp_arima$coef
# \phi_0:
(1-sum(coefs[1:gnp_ar$order])) * coefs[['intercept']]
# Residual standard error:
sqrt(gnp_arima$sigma2)
# Same logic applies when calculate cycles
```

### Stationary


## Identifying AR Models in Practice

### Partial Autocorrelation Function (PACF)

```{r}
df = read.table(dataset$m_ibm3dx2608$url, header=T)
crsp = df$vwrtn
crsp_pacf = pacf(crsp)
crsp_pacf

mdl = arima(crsp, order=c(3,0,0))
mdl
(1-sum(mdl$coef)+mdl$coef['intercept'])*mdl$coef['intercept']  # phi0

rtest = Box.test(mdl$residuals, lag=12, type='Ljung')
rtest

# Adjust degrees of freedom:
1-pchisq(rtest$statistic, 12-3)

# To fix AR(2) coef to zero:
arima(crsp, order=c(3,0,0), fixed=c(NA,0,NA,NA))
```

`Question:`{.text-warning} How to get p-values of coeffcients?

`Answer:`{.text-success} A general form is 
`pt(abs(model$coef)/sqrt(diag(aa$var.coef)), df.of.resid, lower.tail = F)*2`, 
but you need to find out how to access to 
coef and stardard of coefs of your model object.

### Information Criteria

You can use `ar(...)$aic` and `ar(...)$order` to get IC based
order selection data.


## Goodness of Fit


## Forecasting

See the following code example.

```{r}
dt = read.table(dataset$m_ibm3dx2608$url, header=T)
vwval = dt$vwrtn
vw = ts(vwval, frequency=12, start=c(1926,1))
x = ts(vw[1:984], start=start(vw), frequency=12)
actual = ts(vw[-c(1:984)], end=end(vw), frequency=12)
fit = forecast::Arima(x, order=c(3,0,0))
fore1 = forecast::forecast(fit, 12)
plot(x, xlim=c(2000, 2010))
points(fore1$mean, pch=1, cex=0.8, type="o")
points(actual, pch=20, cex=0.8, type="o")
lines(ts(fore1$upper[,'95%'], end=end(vw), frequency=12), lty=3)
lines(ts(fore1$lower[,'95%'], end=end(vw), frequency=12), lty=3)
```

To produce exactly the same figure as Figure 2.7 on page57:

```{r}
x = ts(vw[1:984], start=start(vw), frequency=12)
actual = ts(vw[-c(1:984)], end=end(vw), frequency=12)
fit = forecast::Arima(x, order=c(3,0,0))
fore1 = forecast::forecast(fit, 12)
plot(vw, xlim=c(2000, 2010), type="o", pch=20, cex=0.5)
points(fore1$mean, pch=1, cex=0.8, type="o")
points(actual, pch=20, cex=0.8, type="o")
lines(ts(c(x, fore1$mean), frequency=12, start=start(x)))
x_upper = ts(c(x, fore1$upper[,'95%']), start=start(x), frequency=12)
x_lower = ts(c(x, fore1$lower[,'95%']), start=start(x), frequency=12)
lines(x_upper, lty=3)
lines(x_lower, lty=3)
```


# MA Models

$$ r_t = c_0 + a_t - \theta_1 a_{t-1} - /ldots - \theta_q a_{t-2} $$

## Properties of MA models

### Stationarity

$$ E(r_t) = c_0 $$
$$ \text{Var}(r_t) = (1+\theta_1^2 + \theta_2^2 + \ldots + \theta_q^2)\sigma_a^2 $$

### Autocorrelation Function

For an MA(q) model, $\rho_\ell = 0, for \ell > q $.

### Invertibility

Rewriting a zero-mean MA(1) mdel as $ a_t = r_t + \theta_1 a_{t-1} $,
$$ a_t = r_t + \theta_1 r_{t-1} + \theta_1^2 r_{t-2} + \ldots $$

if $ |\theta_1| < 1$, we say the MA(1) model is **invertible**.

## Identifying MA Order

```{r}
dt = read.table(dataset$m_ibm3dx2608$url, header=T)
ew = dt$ewrtn
ew = ts(ew, frequency=12, start=c(1926,1))
plot(ew, type='l')
ew_acf = acf(ew)
ew_acf
```

## Estimating an MA model

```{r}
m1 = arima(ew, order=c(0,0,9), fixed=c(NA,0,NA,rep(0,5),NA, NA))
m1
```







# Simple ARMA Models

An ARMA(1,1) Model
$$ r_t - \phi_1 r_{t-1} = \phi_0 + a_t - \theta_1 a_{t-1} $$

## Properties of ARMA(1,1) Models

$$ E(r_t) = \mu = \frac{\phi_1}{1-\phi_1} $$

$$ \text{Var}(r_t) = 
\frac{(1-2\phi_1 \theta_1 + \theta_1^2)\sigma_a^2}{1-\phi_1^2} $$

# Unitroot, Seasonal Models etc., to be finished

TO BE FINISHED...

```{r}
timeDate::setRmetricsOptions(myFinCenter = "GMT")
charvec = timeDate::timeCalendar()
set.seed(4711)
data = matrix(exp(cumsum(rnorm(12, sd = 0.1))))
tS = timeSeries::timeSeries(data, charvec, units = "tS")
tS

```

```{r unitroot}
da = read.table(dataset$d_sp55008$url, header=T)
sp5 = log(da[,7])
fUnitRoots::adfTest(sp5, lags=15, type=c('ct'))
```


Forecasting an ARIMA Model

```{r}
data(gtemp, package="astsa")
fit1 = arima(gtemp, order=c(1,1,1))   
fore1 = predict(fit1, 15)   
nobs = length(gtemp)
fit2 = arima(gtemp, order=c(1,1,1), xreg=1:nobs)  
fore2 = predict(fit2, 15, newxreg=(nobs+1):(nobs+15))
ok=par()
par(mfrow=c(2,1))
ts.plot(gtemp,fore1$pred, col=1:2, main="WRONG")
ts.plot(gtemp,fore2$pred, col=1:2, main="RIGHT")

fit3 = forecast::Arima(gtemp, order=c(1,1,1),include.drift=T)
fore3 = forecast::forecast(fit3, 15)
ts.plot(gtemp,fore2$pred, col=1:2, main="RIGHT")
ts.plot(gtemp,fore3$mean, col=1:2, main="RIGHT")
```